---
title: "R Functions"
author: "Kajola Gbenga, Data Scientist & AI Engineer"
output: html_document
---

# R Functions

A function is a block of code which only runs when it is called.

You can pass data, known as arguments, into a function.

A function can return data as a result using the `return()` statement.

## Creating a Function

In R, a function is defined using the `function` keyword, assigned to a name with `<-`.

## Defining a Function

Syntax:

```r
function_name <- function(parameters) {
  # block of code
}
```

This is a well detailed explanation of the different components in the syntax above when creating a function:

* `function_name` - the name of your function (follow standard naming conventions).
* `parameters` - inputs (optional) the function needs to work with.
* comments (`#`) - used instead of a docstring, to briefly describe what the function does.
* `{ }` block - the code that the function executes, enclosed in curly braces.

```{r}
# e.g

my_function <- function() {
  print("Hello from a function")
}

my_function()
```

## Calling a Function

After defining it, you call a function by writing its name followed by parentheses `()`.

In other words, to successfully call a function, use the function name followed by parentheses:

```{r}
# e.g

my_function <- function() {
  # I created this function to say HI from the function.
  print("Hello from a function")
}

my_function()
```

```{r}
# another example

greet <- function() {
  print("Hello, welcome to R!")
}

greet()  # Function call
```

```{r}
greet_user <- function(name) {
  print(paste0("Hello, ", name, "! Welcome to R...."))
}

greet_user("Jennifer")
```

## Parameters and Arguments

Parameter: A variable listed inside the parentheses in the function definition.

Argument: A value sent to the function when it is called.

## Arguments

Information can be passed into functions as arguments.

Arguments are specified after the function name, inside the parentheses. You can add as many arguments as you want, just separate them with a comma.

The example below has a function with one argument (`first_name`). When the function is called, we pass along a first name, which is used inside the function to print the full name:

```{r}
my_function <- function(first_name) {
  print(paste(" Hi", first_name))
}

my_function("Comfort")
my_function("Tata")
my_function("Ajanlekoko")
```

```{r}
my_function("Gbenga")
```

```{r}
greet_user <- function(name) {  # 'name' is the parameter
  print(paste0("Thank you ", name, " for investing with us!"))
}

greet_user("Comfort")  # 'Comfort' is the argument
```

You can have multiple parameters, separated by commas.

```{r}
multiply_numbers <- function(a, b) { # Creating a function called multiply_numbers which allows two parameters
  return(a * b)                       # The output should be the multiplication of the two parameters
}

# Now to use the function
result <- multiply_numbers(15, 30) # Enters the two required arguments to be passed into the function
print(result)
```

## Number of Arguments

By default, a function must be called with the correct number of arguments. Meaning that if your function expects 2 arguments, you have to call the function with 2 arguments, not more, and not less.

```{r}
# In the example below, this function expects 2 arguments, and gets 2 arguments:

my_function <- function(fname, lname) {
  print(paste(fname, lname))
}

my_function("Comfort", "Ajanlekoko")
```

NOTE => if you try to call the function created above with 1 or 3 arguments, you will get an error as shown in the example below.
This is because while creating the function, two arguments were defined with it.
So for an error not to be raised while calling the function, the number of parameters defined must match the number of arguments used when it is called, unless a default value is set.

```{r, eval=FALSE}
# This function expects 2 arguments, but gets only 1:

my_function <- function(fname, lname) {
  print(paste(fname, lname))
}

my_function("Emil")
```

# BAR CHARTS WITHOUT USING FUNCTIONS

```{r}
# BAR CHARTS WITHOUT USING FUNCTIONS

# My data
students <- c("Ade", "Bola", "Chidi", "Zainab")
scores <- c(78, 85, 69, 90)

# My plot
barplot(scores,
        names.arg = students,
        main = "Student Scores",
        xlab = "Students",
        ylab = "Scores")
```

## Bar Chart Using a Function

```{r}
# My data
students <- c("Ade", "Bola", "Chidi", "Zainab")
scores <- c(78, 85, 69, 90)

Gbefun <- function(students, scores, title, xlabel, ylabel) {
  barplot(scores,
          names.arg = students,
          main = title,
          xlab = xlabel,
          ylab = ylabel)
}

Gbefun(students, scores, "Student Scores", "Students", "Scores")
```
