---
title: "Project 1: Nigerian Retail Sales EDA"
subtitle: "Exploratory Data Analysis with the Tidyverse"
author: "Prodigy Training Hub"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 10,
  fig.height = 6
)
```

# Cell 1: Load Libraries

```{r load-libraries}
library(tidyverse)   # Core: dplyr, ggplot2, tidyr, readr, stringr, purrr, forcats
library(janitor)     # clean_names(), tabyl()
library(skimr)       # skim() - rich summary statistics
library(lubridate)   # Date parsing and manipulation
library(scales)      # Number formatting (comma, naira, percent)
library(knitr)       # kable() for clean table output
```

---

# Cell 2: Create the Dataset

```{r create-dataset}
set.seed(42)
n <- 1500

states <- c("Lagos", "Abuja", "Kano", "Rivers", "Oyo",
            "Delta", "Anambra", "Kaduna", "Enugu", "Ondo")

regions <- c("South West", "North Central", "North West", "South South", "South West",
             "South South", "South East", "North West", "South East", "South West")

state_region <- tibble(state = states, region = regions)

categories <- c("Electronics", "Fashion & Apparel", "Food & Beverages",
                "Home & Kitchen", "Health & Beauty", "Agriculture & Farming")

payment_methods <- c("Cash", "POS", "Bank Transfer", "Mobile Money")

raw_retail <- tibble(
  transaction_id   = paste0("TXN-", str_pad(1:n, 5, pad = "0")),
  date             = sample(seq(as.Date("2023-01-01"), as.Date("2023-12-31"), by = "day"), n, replace = TRUE),
  state            = sample(states, n, replace = TRUE, prob = c(0.25, 0.15, 0.12, 0.10, 0.09, 0.08, 0.07, 0.06, 0.05, 0.03)),
  store_id         = paste0("STR-", sample(sprintf("%03d", 1:40), n, replace = TRUE)),
  category         = sample(categories, n, replace = TRUE, prob = c(0.20, 0.22, 0.18, 0.15, 0.13, 0.12)),
  units_sold       = sample(1:50, n, replace = TRUE),
  unit_price_ngn   = round(runif(n, 500, 150000), -1),
  discount_pct     = sample(c(0, 0, 0, 5, 10, 15, 20), n, replace = TRUE),
  payment_method   = sample(payment_methods, n, replace = TRUE, prob = c(0.30, 0.35, 0.20, 0.15)),
  customer_age     = sample(18:65, n, replace = TRUE),
  customer_gender  = sample(c("Male", "Female"), n, replace = TRUE, prob = c(0.54, 0.46)),
  returns          = sample(c(0, 1), n, replace = TRUE, prob = c(0.92, 0.08))
) |>
  # Inject missing values realistically
  mutate(
    customer_age    = ifelse(runif(n) < 0.04, NA, customer_age),
    discount_pct    = ifelse(runif(n) < 0.02, NA, discount_pct),
    customer_gender = ifelse(runif(n) < 0.01, NA, customer_gender)
  )

cat("Dataset created:", nrow(raw_retail), "rows x", ncol(raw_retail), "columns\n")
```

---

# Cell 3: First Look at the Data

```{r first-look}
glimpse(raw_retail)
```

---

# Cell 4: View First 10 Rows

```{r head}
head(raw_retail, 10)
```

---

# Cell 5: Rich Summary with skimr

```{r skim-summary}
skim(raw_retail)
```

---

# Cell 6: Check Missing Values

```{r missing-values}
missing_summary <- raw_retail |>
  summarise(across(everything(), ~ sum(is.na(.)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "missing_count") |>
  mutate(
    missing_pct = round(missing_count / nrow(raw_retail) * 100, 2),
    has_missing = missing_count > 0
  ) |>
  arrange(desc(missing_count))

missing_summary |> kable(caption = "Missing Value Summary")
```

---

# Cell 7: Clean Column Names

```{r clean-names}
retail <- raw_retail |>
  clean_names()

names(retail)
```

---

# Cell 8: Engineer Derived Columns

```{r feature-engineering}
retail <- retail |>
  left_join(state_region, by = "state") |>
  mutate(
    revenue_ngn     = units_sold * unit_price_ngn * (1 - discount_pct / 100),
    revenue_ngn     = replace_na(revenue_ngn, units_sold * unit_price_ngn),
    month           = month(date, label = TRUE, abbr = TRUE),
    quarter         = paste0("Q", quarter(date)),
    week_number     = week(date),
    day_of_week     = wday(date, label = TRUE, abbr = FALSE),
    is_weekend      = day_of_week %in% c("Saturday", "Sunday"),
    price_tier      = case_when(
      unit_price_ngn < 5000              ~ "Budget",
      unit_price_ngn < 30000             ~ "Mid-Range",
      unit_price_ngn < 80000             ~ "Premium",
      TRUE                               ~ "Luxury"
    ),
    age_group       = case_when(
      customer_age < 25                  ~ "18-24",
      customer_age < 35                  ~ "25-34",
      customer_age < 45                  ~ "35-44",
      customer_age < 55                  ~ "45-54",
      TRUE                               ~ "55+"
    ),
    discount_applied = !is.na(discount_pct) & discount_pct > 0
  )

glimpse(retail)
```

---

# Cell 9: Handle Missing Values

```{r handle-missing}
retail <- retail |>
  mutate(
    customer_age    = ifelse(is.na(customer_age), median(customer_age, na.rm = TRUE), customer_age),
    discount_pct    = ifelse(is.na(discount_pct), 0, discount_pct),
    customer_gender = ifelse(is.na(customer_gender), "Unknown", customer_gender)
  )

cat("Missing values remaining:", sum(is.na(retail)), "\n")
```

---

# Cell 10: Revenue Summary Statistics

```{r revenue-summary}
retail |>
  summarise(
    total_transactions = n(),
    total_revenue_ngn  = sum(revenue_ngn),
    avg_revenue_ngn    = mean(revenue_ngn),
    median_revenue_ngn = median(revenue_ngn),
    min_revenue_ngn    = min(revenue_ngn),
    max_revenue_ngn    = max(revenue_ngn),
    total_units_sold   = sum(units_sold),
    total_returns      = sum(returns),
    return_rate_pct    = round(sum(returns) / n() * 100, 2)
  ) |>
  mutate(across(ends_with("_ngn"), ~ comma(.))) |>
  kable(caption = "Overall Revenue Summary")
```

---

# Cell 11: Revenue by State

```{r revenue-by-state}
revenue_by_state <- retail |>
  group_by(state, region) |>
  summarise(
    transactions       = n(),
    total_revenue_ngn  = sum(revenue_ngn),
    avg_revenue_ngn    = mean(revenue_ngn),
    total_units        = sum(units_sold),
    .groups = "drop"
  ) |>
  arrange(desc(total_revenue_ngn))

revenue_by_state |>
  mutate(across(ends_with("_ngn"), ~ paste0("N", comma(round(., 0))))) |>
  kable(caption = "Revenue by State")
```

---

# Cell 12: Revenue by Product Category

```{r revenue-by-category}
revenue_by_cat <- retail |>
  group_by(category) |>
  summarise(
    transactions      = n(),
    total_revenue_ngn = sum(revenue_ngn),
    avg_unit_price    = mean(unit_price_ngn),
    avg_units_sold    = mean(units_sold),
    return_rate_pct   = round(sum(returns) / n() * 100, 2),
    .groups = "drop"
  ) |>
  arrange(desc(total_revenue_ngn))

revenue_by_cat |>
  mutate(across(ends_with("_ngn") | starts_with("avg_unit"), ~ paste0("N", comma(round(., 0))))) |>
  kable(caption = "Revenue by Product Category")
```

---

# Cell 13: Monthly Revenue Trend

```{r monthly-trend}
monthly_revenue <- retail |>
  group_by(month, quarter) |>
  summarise(
    total_revenue_ngn = sum(revenue_ngn),
    transactions      = n(),
    .groups = "drop"
  )

monthly_revenue |>
  mutate(total_revenue_ngn = paste0("N", comma(round(total_revenue_ngn, 0)))) |>
  kable(caption = "Monthly Revenue Summary")
```

---

# Cell 14: Payment Method Analysis

```{r payment-analysis}
retail |>
  group_by(payment_method) |>
  summarise(
    count          = n(),
    share_pct      = round(n() / nrow(retail) * 100, 2),
    avg_revenue    = round(mean(revenue_ngn), 0),
    total_revenue  = sum(revenue_ngn),
    .groups = "drop"
  ) |>
  arrange(desc(count)) |>
  kable(caption = "Payment Method Breakdown")
```

---

# Cell 15: Customer Demographics Analysis

```{r demographics}
retail |>
  filter(customer_gender != "Unknown") |>
  group_by(customer_gender, age_group) |>
  summarise(
    count          = n(),
    avg_revenue    = round(mean(revenue_ngn), 0),
    total_revenue  = round(sum(revenue_ngn), 0),
    .groups = "drop"
  ) |>
  arrange(customer_gender, age_group) |>
  kable(caption = "Revenue by Gender and Age Group")
```

---

# Cell 16: Discount Impact Analysis

```{r discount-analysis}
retail |>
  group_by(discount_applied, discount_pct) |>
  summarise(
    transactions   = n(),
    avg_units_sold = round(mean(units_sold), 2),
    avg_revenue    = round(mean(revenue_ngn), 0),
    return_rate    = round(mean(returns) * 100, 2),
    .groups = "drop"
  ) |>
  filter(!(discount_applied == FALSE & discount_pct > 0)) |>
  kable(caption = "Discount vs Revenue and Returns")
```

---

# Cell 17: Weekend vs Weekday Sales

```{r weekend-analysis}
retail |>
  group_by(is_weekend) |>
  summarise(
    transactions    = n(),
    avg_revenue_ngn = round(mean(revenue_ngn), 0),
    total_revenue   = round(sum(revenue_ngn), 0),
    avg_units       = round(mean(units_sold), 2),
    .groups = "drop"
  ) |>
  mutate(day_type = ifelse(is_weekend, "Weekend", "Weekday")) |>
  select(day_type, transactions, avg_revenue_ngn, total_revenue, avg_units) |>
  kable(caption = "Weekend vs Weekday Performance")
```

---

# Cell 18: Price Tier Distribution

```{r price-tier}
retail |>
  count(price_tier, category) |>
  pivot_wider(names_from = price_tier, values_from = n, values_fill = 0) |>
  kable(caption = "Transactions by Category and Price Tier")
```

---

# Cell 19: Chart 1 - Revenue by State (Bar Chart)

```{r chart-revenue-state}
retail |>
  group_by(state) |>
  summarise(total_revenue = sum(revenue_ngn), .groups = "drop") |>
  mutate(state = fct_reorder(state, total_revenue)) |>
  ggplot(aes(x = state, y = total_revenue / 1e6, fill = total_revenue)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  scale_fill_gradient(low = "#8ecae6", high = "#023047") +
  scale_y_continuous(labels = function(x) paste0("N", x, "M")) +
  labs(
    title    = "Total Revenue by State (2023)",
    subtitle = "Nigerian Retail Sales Dataset - 1,500 Transactions",
    x        = NULL,
    y        = "Total Revenue (Millions NGN)",
    caption  = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold"),
    plot.subtitle = element_text(colour = "grey50"),
    panel.grid.major.y = element_blank()
  )
```

---

# Cell 20: Chart 2 - Revenue by Category (Lollipop)

```{r chart-category-lollipop}
retail |>
  group_by(category) |>
  summarise(total_revenue = sum(revenue_ngn), .groups = "drop") |>
  mutate(category = fct_reorder(category, total_revenue)) |>
  ggplot(aes(x = category, y = total_revenue / 1e6)) +
  geom_segment(aes(xend = category, yend = 0), colour = "#457b9d", linewidth = 1) +
  geom_point(size = 5, colour = "#e63946") +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0("N", x, "M")) +
  labs(
    title   = "Revenue by Product Category (2023)",
    x       = NULL,
    y       = "Total Revenue (Millions NGN)",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title         = element_text(face = "bold"),
    panel.grid.major.y = element_blank()
  )
```

---

# Cell 21: Chart 3 - Monthly Revenue Trend (Line Chart)

```{r chart-monthly-trend}
monthly_revenue |>
  ggplot(aes(x = as.integer(month), y = total_revenue_ngn / 1e6, group = 1)) +
  geom_line(colour = "#2a9d8f", linewidth = 1.2) +
  geom_point(aes(colour = quarter), size = 4) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_y_continuous(labels = function(x) paste0("N", x, "M")) +
  scale_colour_manual(values = c(Q1 = "#264653", Q2 = "#2a9d8f", Q3 = "#e9c46a", Q4 = "#e76f51")) +
  labs(
    title   = "Monthly Revenue Trend (2023)",
    x       = NULL,
    y       = "Revenue (Millions NGN)",
    colour  = "Quarter",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Cell 22: Chart 4 - Payment Method Distribution (Donut)

```{r chart-payment-donut}
payment_data <- retail |>
  count(payment_method) |>
  mutate(
    pct   = n / sum(n),
    label = paste0(payment_method, "\n", percent(pct, accuracy = 0.1))
  )

ggplot(payment_data, aes(x = 2, y = pct, fill = payment_method)) +
  geom_col(width = 1, colour = "white", linewidth = 0.5) +
  coord_polar(theta = "y") +
  xlim(0.5, 2.5) +
  scale_fill_manual(values = c(
    "Cash"           = "#264653",
    "POS"            = "#2a9d8f",
    "Bank Transfer"  = "#e9c46a",
    "Mobile Money"   = "#e76f51"
  )) +
  geom_text(aes(label = label), position = position_stack(vjust = 0.5),
            colour = "white", size = 3.5, fontface = "bold") +
  labs(
    title   = "Payment Method Distribution (2023)",
    fill    = NULL,
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_void(base_size = 13) +
  theme(
    plot.title  = element_text(face = "bold", hjust = 0.5),
    legend.position = "none"
  )
```

---

# Cell 23: Chart 5 - Revenue by Region and Category (Faceted Bar)

```{r chart-region-category-facet}
retail |>
  group_by(region, category) |>
  summarise(total_revenue = sum(revenue_ngn) / 1e6, .groups = "drop") |>
  mutate(category = str_wrap(category, width = 12)) |>
  ggplot(aes(x = category, y = total_revenue, fill = region)) +
  geom_col(position = "dodge") +
  facet_wrap(~ region, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = function(x) paste0("N", x, "M")) +
  labs(
    title   = "Revenue by Category across Geopolitical Regions (2023)",
    x       = NULL,
    y       = "Revenue (Millions NGN)",
    fill    = "Region",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title  = element_text(face = "bold"),
    legend.position = "none",
    axis.text.x = element_text(angle = 35, hjust = 1, size = 8)
  )
```

---

# Cell 24: Chart 6 - Unit Price Distribution by Category (Boxplot)

```{r chart-boxplot-price}
retail |>
  mutate(category = fct_reorder(category, unit_price_ngn, median)) |>
  ggplot(aes(x = category, y = unit_price_ngn / 1000, fill = category)) +
  geom_boxplot(alpha = 0.7, outlier.colour = "#e63946", outlier.alpha = 0.4) +
  coord_flip() +
  scale_fill_brewer(palette = "Pastel1") +
  scale_y_continuous(labels = function(x) paste0("N", x, "K")) +
  labs(
    title   = "Unit Price Distribution by Product Category",
    x       = NULL,
    y       = "Unit Price (Thousands NGN)",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "none"
  )
```

---

# Cell 25: Chart 7 - Revenue Heatmap (Month x Day of Week)

```{r chart-heatmap}
retail |>
  group_by(month, day_of_week) |>
  summarise(avg_revenue = mean(revenue_ngn) / 1000, .groups = "drop") |>
  mutate(day_of_week = factor(day_of_week,
    levels = c("Monday","Tuesday","Wednesday","Thursday","Friday","Saturday","Sunday"))) |>
  ggplot(aes(x = month, y = day_of_week, fill = avg_revenue)) +
  geom_tile(colour = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#caf0f8", high = "#023e8a",
                      labels = function(x) paste0("N", x, "K")) +
  labs(
    title   = "Average Transaction Value Heatmap (Month x Day of Week)",
    x       = NULL,
    y       = NULL,
    fill    = "Avg Revenue",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title  = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

---

# Cell 26: Chart 8 - Age Group x Gender Revenue (Stacked Bar)

```{r chart-demographics-stacked}
retail |>
  filter(customer_gender != "Unknown") |>
  group_by(age_group, customer_gender) |>
  summarise(total_revenue = sum(revenue_ngn) / 1e6, .groups = "drop") |>
  ggplot(aes(x = age_group, y = total_revenue, fill = customer_gender)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(Male = "#457b9d", Female = "#e63946")) +
  scale_y_continuous(labels = function(x) paste0("N", x, "M")) +
  labs(
    title   = "Total Revenue by Age Group and Gender",
    x       = "Age Group",
    y       = "Revenue (Millions NGN)",
    fill    = "Gender",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Cell 27: Chart 9 - Scatter Plot - Units Sold vs Revenue

```{r chart-scatter}
retail |>
  sample_n(500) |>
  ggplot(aes(x = units_sold, y = revenue_ngn / 1000, colour = category)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8) +
  facet_wrap(~ category, scales = "free_y") +
  scale_colour_brewer(palette = "Dark2") +
  scale_y_continuous(labels = function(x) paste0("N", x, "K")) +
  labs(
    title   = "Units Sold vs Revenue by Category",
    x       = "Units Sold",
    y       = "Revenue (Thousands NGN)",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    plot.title      = element_text(face = "bold"),
    legend.position = "none"
  )
```

---

# Cell 28: Chart 10 - Quarterly Revenue by Region (Grouped Bar)

```{r chart-quarterly-region}
retail |>
  group_by(quarter, region) |>
  summarise(total_revenue = sum(revenue_ngn) / 1e6, .groups = "drop") |>
  ggplot(aes(x = quarter, y = total_revenue, fill = region)) +
  geom_col(position = "dodge") +
  scale_fill_brewer(palette = "Set1") +
  scale_y_continuous(labels = function(x) paste0("N", x, "M")) +
  labs(
    title   = "Quarterly Revenue by Geopolitical Region",
    x       = "Quarter",
    y       = "Revenue (Millions NGN)",
    fill    = "Region",
    caption = "Source: NigeriaRetail Co. Dataset | Prodigy Training Hub"
  ) +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))
```

---

# Cell 29: Top 10 Stores by Revenue

```{r top-stores}
retail |>
  group_by(store_id, state) |>
  summarise(
    transactions  = n(),
    total_revenue = sum(revenue_ngn),
    avg_basket    = mean(revenue_ngn),
    .groups = "drop"
  ) |>
  slice_max(order_by = total_revenue, n = 10) |>
  mutate(total_revenue = paste0("N", comma(round(total_revenue, 0))),
         avg_basket    = paste0("N", comma(round(avg_basket, 0)))) |>
  kable(caption = "Top 10 Stores by Total Revenue")
```

---

# Cell 30: Return Rate Analysis by Category and Region

```{r return-analysis}
retail |>
  group_by(region, category) |>
  summarise(
    transactions = n(),
    returns      = sum(returns),
    return_rate  = round(mean(returns) * 100, 2),
    .groups = "drop"
  ) |>
  pivot_wider(names_from = category, values_from = return_rate, values_fill = 0) |>
  kable(caption = "Return Rate (%) by Region and Category")
```

---

# Cell 31: Correlation - Numeric Variables

```{r correlation}
retail |>
  select(units_sold, unit_price_ngn, discount_pct, revenue_ngn, customer_age) |>
  cor(use = "complete.obs") |>
  round(3) |>
  kable(caption = "Correlation Matrix - Numeric Variables")
```

---

# Cell 32: Key Insights Summary

```{r insights-summary}
top_state    <- retail |> group_by(state) |> summarise(rev = sum(revenue_ngn)) |> slice_max(rev, n=1) |> pull(state)
top_cat      <- retail |> group_by(category) |> summarise(rev = sum(revenue_ngn)) |> slice_max(rev, n=1) |> pull(category)
top_payment  <- retail |> count(payment_method) |> slice_max(n, n=1) |> pull(payment_method)
best_quarter <- retail |> group_by(quarter) |> summarise(rev = sum(revenue_ngn)) |> slice_max(rev, n=1) |> pull(quarter)
total_rev    <- sum(retail$revenue_ngn)
return_rate  <- mean(retail$returns) * 100

cat("========================================\n")
cat("  NIGERIAN RETAIL EDA - KEY INSIGHTS\n")
cat("========================================\n")
cat("Total Revenue:         N", comma(round(total_rev, 0)), "\n")
cat("Total Transactions:   ", nrow(retail), "\n")
cat("Top State:            ", top_state, "\n")
cat("Top Category:         ", top_cat, "\n")
cat("Preferred Payment:    ", top_payment, "\n")
cat("Best Quarter:         ", best_quarter, "\n")
cat("Overall Return Rate:  ", round(return_rate, 2), "%\n")
cat("========================================\n")
```
